Synthesizing Quotient Lenses

### 2017-10-18 papers/QuotientLensSynth/main.tex
Massive changes to intro

\section{Introduction}
\ifplentyoftime
\bcp{Not sure there is time to fix this, but the intro seems very long and
  technical---it takes a long time to get to saying what our actual
  contributions are!  Wonder if we should move most of it to a second
  ``Example'' section.} 
\fi
\bcp{This sentence could be clearer!}{\em Bidirectional programming languages} such as
Boomerang~\cite{boomerang}, Augeas~\cite{augeas}, XSugar~\cite{xsugar},
biXid~\cite{bixid} and X/Inv~\cite{xinv}\bcp{there are many others; we
  should mention all of them} enable a programmer to express a
transformation that maps some input data to some output data, as well as a
transformation that maps the output data back into the input data in a single
expression.  An example of a {\em bidirectional transformation} which we shall
revisit severally in this paper is a program which converts citation records
such as
\begin{verbatim}
@Book {conway,
  Author = "J. H. Conway",
  Title = {Regular Algebra and Finite Machines}
}
\end{verbatim}

@@

Bidirectional programs have also been called {\em lenses}. Formally, Foster et
al~\cite{foster} defined a lens $\ell : S \Leftrightarrow V$ that maps between
an ``abstract" {\em source} $S$ and a ``concrete" {\em view} $V$ to be a triple
of functions $(\ell.\get, \; \ell.\lput, \; \ell.\create)$ satisfying the
following equations:\bcp{Why are we showing all the complexity of asymmetric
lenses?  We should just say that we're considering only the bijective case
for now and then just give the details of that.  (Related: The
``abstract/concrete'' terminology doesn't make much sense in the bijective
setting.)} 
\begin{align*}
\ell.get &: S \longrightarrow V & \ell.\lput \; (\ell.\get \; s) \; s &= s\\
\ell.put &: V \longrightarrow S \longrightarrow S & \ell.\get \; (\ell.\lput \;
v \; s) &= v\\
\ell.create &: S \longrightarrow V & \ell.\lput \; (\ell.\create \; v) &= v
\end{align*}

The $\get$ function presents a view of the data from the source, the
$\lput$ function enables an updated view to be folded back into the source, and
the $\create$ component produces a copy of the source from a view, using
defaults to fill in information that may have been discarded by the $\get$
function in producing the view. The first law states that the $\lput$ function
must restore all the information discarded by $\get$ when its arguments are an
abstract structure and a concrete structure that generates the very same
abstract structure. The second and third laws state that $\lput$ and $\create$
must propagate all of the information contained in their abstract arguments to
the concrete structure they produce. These laws express fundamental expectations
about how the components of a lens should work together.

However, the constraints that are imposed by the lens laws are sometimes too
strict. For instance, the programmer may wish to ignore updates to the data
that only change the order of fields, or add whitespace characters that in
reality are not part of the data. Foster {\em et al.\relax} therefore proposed
\textit{quotient lenses} ~\cite{quotient} as a solution to this problem.
Quotient lenses, which are implemented as a refinement of the bidirectional
string processing language Boomerang, enable a programmer to write lenses which
transform data that is quotiented out by an equivalence relation, so that
records which are equivalent to each other modulo this relation are considered
the same by the underlying lens.

Formally, given equivalence relations $\sim_S$ and $\sim_V$ defined on the
source $S$ and the view $V$ respectively, a quotient lens $q : S/{\sim_S}
\Leftrightarrow V/{\sim_V}$ is a triple of functions $(\ell.\get, \; \ell.\lput,
\; \ell.\create)$ which satisfy the lens laws up to equivalence:
\begin{align*}
\ell.get &: S \longrightarrow V & \ell.\lput \; (\ell.\get \; s) \; s &\sim_S
s\\
\ell.put &: V \longrightarrow S \longrightarrow S & \ell.\get \; (\ell.\lput \;
v \; s) &\sim_V v\\
\ell.create &: S \longrightarrow V & \ell.\lput \; (\ell.\create \; v) &\sim_V v
\end{align*}
Moreover, each of the $\get$, $\lput$ and $\create$ functions must respect
$\sim_S$ and $\sim_V$. That is, if $s \sim_S s'$ then $\get \; s \sim_V
\get \; s'$, if $s \sim_S s'$ and $v \sim_V v'$ then $\lput \; v \; s \sim_S
\lput \; v' \; s'$ and if $v \sim_V v'$, then $\create \; v \sim_S \create \;
v'$.

\bcp{I'm nervous about making the concept of canonizer so central, since it
  is not going to be much of a focus for the rest of the paper.  In general,
it feels like the intro is explaining too much.  We need to find a more
streamlined path through the ideas!}
In Boomerang, equivalence relations are expressed using {\em canonizers}
which are functions that map members of a regular language to their canonical
representation modulo an equivalence relation. Therefore, in order to give a
specification for the data, the programmer often has to write a
regular expression which matches all possible representations of the data, and
then provide a canonizer that does the job of quotienting the data.

For example, suppose that the programmer wants to
derive the \bibtex{} to EndNote transformation in such a way that the program
allows for whitespace to occur arbitrarily between the words in the author and
title fields, ignores the order in which the author and title fields occur in
each record, and allows for arbitrary amounts of whitespace in between the
author and title fields. These rules allow for the \bibtex{} citations
\begin{verbatim}
    @Book {conway,
      Author = "J. H. Conway",
      Title = {Regular Algebra and Finite Machines},
    }
\end{verbatim}
and
\begin{verbatim}
    @Book{     conway,
        Title = {Regular Algebra 
                 and Finite Machines},
           Author = "J. H. Conway",     }
\end{verbatim}
to be considered the same (with the EndNote records having a similar
equivalence relation defined on them).

\begin{figure}[t]
\begin{lstlisting}
let project (r : regexp) (s : string) = canonizer_of_lens ((del r) . (ins s))
let wsp : regexp = [ \n\t\r]$^+$
let wsp_to_space : canonizer = project wsp " "
let valid_name : canonizer = 
    name | (name . wsp_to_space . (name .wsp_to_space)$^*$ .name)
let wsp_to_and : canonizer = project (wsp . "and" . wsp) " and "
let bib_names : canonizer =  
    valid_name | (valid_name . wsp_to_and . (valid_name . wsp_to_and)$^*$ . valid_name)
let bib_author : canonizer = "author = \"" . bib_names . "\""
let label : regexp = [a-zA-Z0-9]$^+$
let title : canonizer = 
    word | (word . wsp_to_space . (word . wsp_to_space)$^*$ . word)
let bib_title : canonizer = "title = {" . title . "}"
let wsp_to_comma : canonizer = project ("," . wsp) ",\n"
let first_perm : canonizer = bib_author . wsp_to_comma . bib_title
let second_perm : canonizer = bib_title . wsp_to_comma . bib_author
let bibtex : canonizer = 
    "@book{" . label. wsp_to_comma . (first_perm | second_perm) . "}"
\end{lstlisting}

  \caption{Boomerang canonizers for the \bibtex{}--EndNote equivalence
    relations.}
  \label{fig:example-canonizers}
\end{figure}
The code sample in Figure~\ref{fig:example-canonizers} shows how the programmer
might specify these equivalence relations on the \bibtex{} records in Boomerang
using canonizers.
%
Using the in-built \cd{canonizer_of_lens} function which extracts a canonizer
from a lens, it is easy to define an equivalence relation that sends all
members of a regular language to a canonical representative.\bcp{Oh dear.
  The \cd{canonizer_of_lens} thing is really kind of a hack---we're using
  the fancy/specialized language of lenses to write what could really be a
  simple program in an arbitrary language (and then throwing away half of the
  lens, when the whole reason we give for having lenses in the first place
  is because we want both directions).  As an implementation trick, it was a
good idea.  But trying to get readers to understand it here does not seem
like a good idea.} For example
the expression 
\begin{lstlisting}
let wsp_to_space : canonizer = canonizer_of_lens ((del wsp) . (ins " "))
\end{lstlisting}
\noindent defines a canonizer which sends arbitrary occurences of whitespace
characters to a single space\bcp{Going into this level of detail encourages
readers to actually try to understand the details, but we don't actually
give enough details to allow them to do that!}. However, a problem arises
when trying to specify 
the equivalence relation which nests this projection relation with the relation
that considers all permutations of the \bibtex{} fields equivalent.

Specifically, it becomes necessary to specify each of the permutations
seperately, since Boomerang's in-built permutation combinator does not account
for separators in between the regular expressions being permuted:\bcp{Why
  couldn't we just enrich Boomerang's permutation combinator to do this?}

\begin{lstlisting}
let first_perm : canonizer = bib_author . wsp_to_comma . bib_title
let second_perm : canonizer = bib_title . wsp_to_comma . bib_author
\end{lstlisting}

One possible solution is to permute the alternative canonizers
\cd{bib_author.wsp_to_comma} and \cd{bib_title.wsp_to_comma},
but Boomerang's type-checker will reject this expression since the permutation
combinator accepts regular expressions rather than canonizers\bcp{It feels
  like we're lost, here, in details of exactly what Boomerang does and
  doesn't do.  This doesn't seem especially relevant to quotient lenses in
  general, or to what we're doing in this paper.}. In any case, the
regular expression
\begin{lstlisting}
bib_author . wsp_to_comma . bib_title . wsp_to_comma | 
bib_title . wsp_to_comma . bib_author . wsp_to_comma
\end{lstlisting}
is not equivalent to,
\begin{lstlisting}
bib_author . wsp_to_comma . bib_title  | 
bib_title . wsp_to_comma . bib_author
\end{lstlisting}
thus the programmer would have to write more lenses and canonizers to correct
this. Therefore, in order to utilize this trick, the programmer might choose to
specify all the regular expressions first (including the whole regular
expressions canonized by the \cd{bib_author} and \cd{bib_title}
canonizers), and then the canonizers separately. Alternatively, he/she may stick
to using canonizers and then liberally make use of the built-in
functions \cd{uncanonized_type} and \cd{canonized_type} which return the
regular expression matching all the data and the regular expression matching
the canonical representatives respectively.

In any case, all of these alternatives lead to verbose code that is
difficult to understand. Moreover, if the \bibtex{} records had 3 or more
fields or even nested permutations, then the programmer would quickly find
themselves in a bind with these workarounds.

To address this issue, we propose the language of {\em Quotient Regular
Expressions} or QREs, which express a broad class of equivalence relations
directly on regular languages using a compact and convenient notation.

With QREs we can specify \bibtex{} records as shown in Figure~\ref{fig:example-qre}.

\begin{figure}[t]
\begin{lstlisting}
let wsp_to_space : canonizer = project [ \n\t\r]$^+$ $\to$ " "
let valid_name : canonizer = 
   name | (name . wsp_to_space . (name . wsp_to_space)$^*$ . name)
let wsp_to_and : canonizer = 
   project ([ \n\t\r]$^+$ . "and" . [ \n\t\r]$^+$) $\to$ " and "
let bib_names : canonizer =  
   valid_name | (valid_name . wsp_to_and . (valid_name . wsp_to_and)$^*$ . valid_name)
let bib_author : canonizer = "author = \"" . bib_names . "\""
let label : regexp = [a-zA-Z0-9]$^+$
let title : canonizer = word | (word . wsp_to_space . (word . wsp_to_space)$^*$ . word)
let bib_title : canonizer = "title = {" . title . "}"
let bib_perm  : canonizer = perm (label, bib_author, bib_title) with
   (project ("," . [ \n\t\r]$^*$ ) $\to$ ",\n")
let bibtex : canonizer = "@book{" . bib_perm . "}"
\end{lstlisting}

\caption{QRE specification of \bibtex{} records.  \bcp{I'm afraid this pair
    of examples is not making the point we intended.  I thought we were
    hoping to just say ``Here's the old Boomerang version; here's the QRE
    version; look how much smaller and simpler the second one is.''  But the
    second one {\em isn't} actually much smaller and, at least on first
    impression, doesn't look simpler.}}
  \label{fig:example-qre}
\end{figure}

The expression \bcp{mangled---did you proofread this?}\cd{project [
  \n\t\r]+ -> " "} is a QRE that matches any 
number of whitespace characters, and then considers all strings of whitespace
characters as equivalent with a single space as the canonical representative. 
Another QRE in the example above is the

@@

\noindent QRE, which specifies that the label, title and author fields can occur
in any order, with the fields separated by a comma followed by arbitrary amounts
of whitespace. The canonical representation is that in which the label, author
and title fields occur in that order, separated by a comma followed a newline.
The \cd{perm} QRE is defined at the level of canonizers which allows for
easy nesting of permutations. Similarly, it is easy to specify any number of
fields within the \cd{perm} QRE.

While QREs kill two birds with one stone by specifying the
data and the equivalence relation on the data in one expression,
he/she\bcp{Awkward: just use ``she'' (or find a way to alternate pronouns,
  if that doesn't create confusion)}
still has to manually program the underlying lens which maps the equivalence
classes of the source data to the equivalence classes of the view data, a task
which is itself tedious and error-prone as the code snippet below
illustrates:\bcp{If we added this to the old-Boomerang figure it might make
  a better case for size and simplicity!}

@@

Our approach is to define a class of {\em QRE lenses} which are the same as
bijective lenses with ``canonizers at the edges'' as in
Figure~\ref{fig:attheedges}. In the forward direction, a QRE lens $q$ uses the
source QRE $c$ to compute a unique representative for the data modulo the
equivalence relation defined by $c$ and then applies the $\get$ function of a
bijective lens $\ell$ to this representative. In the backward direction, $q$
operates similarly, but using the view QRE $c'$ and the $\lput$ function of
$\ell$.

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{attheedges}
  \caption{Quotient Lens with ``Canonizers at the Edges''\sam{This is a
      screenshot from Nate's thesis. Are we allowed to use it?}\bcp{We can
      (if we put ``from [Nate's thesis]'' in the caption), but I don't think
      we want the full complexity of asymmetric lenses here.  We should draw
      the figure again.  (I have the OmniGraffle file for it, so this is
      easy if I know how it should look.)}}
  \label{fig:attheedges}
\end{figure}

>>>>

\section{Introduction}
A {\em lens} is a {\em bidirectional transformation} between a source
type $S$ and a target type $T$ in which the forward transformation and the
backward transformation satisfy a certain set of round-tripping laws (e.g.
invertibility).

An example of a bidirectional transformation which we shall revisit
severally in this paper is a lens which converts citation records such as
\begin{verbatim}
@Book {conway,
  Author = "J. H. Conway",
  Title = {Regular Algebra and Finite Machines}
}
\end{verbatim}

@@

\sam{I need a better introduction\ldots}

Lenses are desirable since they save the programmer from having to write two
seperate transformations rather than one. Also, lenses offer strong
guarantees about how the two transformations interact since lenses are
constrained by lens laws.

However, these constraints are sometimes too strict. For example,  the
programmer may want the \bibtex{} to EndNote transformation to allow for
whitespace to occur arbitrarily between the words in the label, author and
title fields as well as in between them, and ignore the order in which the
author and title fields occur in each record. These rules allow for the
\bibtex{} citations
\begin{verbatim}
    @Book {conway,
      Author = "J. H. Conway",
      Title = {Regular Algebra and Finite Machines},
    }
\end{verbatim}
and
\begin{verbatim}
    @Book{     conway,
        Title = {Regular Algebra 
                 and Finite Machines},
           Author = "J. H. Conway",     }
\end{verbatim}
to be considered the same, with the EndNote records having similar rules
defined on them.

As a solution to this problem, Foster et al introduced {\em quotient lenses},
which are a generalization of lenses in which the source and target data are
considered modulo an equivalence relation. Quotient lenses are
both simpler and more flexible, since whole equivalence classes of data
values can be handled all at once and inessential details like whitespace
and ordering of fields can be ignored.

However many complicated equivalence relations are difficult to express in
Boomerang, the lens programming language introduced by Foster et al which
has quotient lens capability. For instance, while it is possible to express the
equivalence on the \bibtex{} data above in Boomerang, actually doing so is
difficult. The difficulty arises from the fact that the entire equivalence is
composed of three separate equivalences which nest and interact with each other
in a way that makes them difficult to express with the standard tools
available. Moreover, many of the equivalences expressible in Boomerang require
that the programmer first specify the data itself in form of a regular
expression before specifying the equivalence relation on the data.

To address this issue, we propose the language of {\em Quotient Regular
Expressions} or QREs, which express a broad class of equivalence relations
directly on regular languages using a compact and convenient notation. For
example, with QREs we can specify \bibtex{} records as shown in
Figure~\ref{fig:example-qre}.

\begin{figure}[t]
\begin{lstlisting}
let wsp_to_space = project [ \n\t\r]$^+$ $\to$ " "
let valid_name = name | (name . wsp_to_space . (name . wsp_to_space)$^*$ . name)
let wsp_to_and = project ([ \n\t\r]$^+$ . "and" . [ \n\t\r]$^+$) $\to$ " and "
let bib_names = valid_name | (valid_name . wsp_to_and . (valid_name . wsp_to_and)$^*$ . valid_name)
let bib_author = "author = \"" . bib_names . "\""
let title = word | (word . wsp_to_space . (word . wsp_to_space)$^*$ . word)
let bib_title = "title = {" . title . "}"
let bib_perm = perm (label, bib_author, bib_title) with(project ("," . [ \n\t\r]$^*$ ) $\to$ ",\n")
let bibtex = "@book{" . bib_perm . "}"
\end{lstlisting}
\caption{QRE specification of \bibtex{} records. }
  \label{fig:example-qre}
\end{figure}

The expression \cd{project [ \\n\\t\\r]+ -> " "} is a QRE that matches any 
number of whitespace characters, and then considers all strings of whitespace
characters as equivalent with a single space as the canonical representative. 
Another QRE in the example above is the

@@

\noindent QRE, which specifies that the label, title and author fields can occur
in any order, with the fields separated by a comma followed by arbitrary amounts
of whitespace. The canonical representation is that in which the label, author
and title fields occur in that order, separated by a comma followed a newline.
The \cd{perm} QRE allows for easy nesting of permutations. Similarly, it is easy
to specify any number of fields within the \cd{perm} QRE.

While QREs kill two birds with one stone by specifying the
data and the equivalence relation on the data in one expression, the programmer
still has to manually derive the underlying lens which maps the equivalence
classes of the source data to the equivalence classes of the view data, a task
which is itself tedious and error-prone as the code snippet below
illustrates:

@@

Our approach, which follows the design of quotient lenses in Boomerang, is to
define a class of {\em QRE lenses} which are the same as bijective lenses with
``canonizers at the edges'' as in Figure~\ref{fig:attheedges}. In the forward
direction, a QRE lens $q$ uses the source QRE $c$ to compute a unique
representative for the data modulo the equivalence relation defined by $c$ and
then applies the $\get$ function of a bijective lens $\ell$ to this
representative. In the backward direction, $q$ operates similarly, but using
the view QRE $c'$ and the $\lput$ function of $\ell$.

\begin{figure}[t]
  \centering
  \includegraphics[width=\textwidth]{canonizers-outside}
  \caption{Quotient Lens with ``Canonizers at the Edges''\sam{This is a
      screenshot from Nate's thesis. Are we allowed to use it?}\bcp{We can
      (if we put ``from [Nate's thesis]'' in the caption), but I don't think
      we want the full complexity of asymmetric lenses here.  We should draw
      the figure again.  (I have the OmniGraffle file for it, so this is
      easy if I know how it should look.)}\bcp{Done---what do you think?}}
  \label{fig:attheedges}
\end{figure}
